---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(brms)
library(arm)
library(loo)
```

```{r}
# Reading in data
set.seed(1234)
setwd("C:/UMass/730/Project")
heart <- read.csv("heart_failure_clinical_records_dataset.csv")
heartt <- heart[sample(nrow(heart), 100), ]
heartt
```

```{r}
dir.create("output")
```

```{r}
# Full model with test data
fullmodtest <- brm(formula = DEATH_EVENT ~ ., 
                data = heartt,
                family = bernoulli(link = "logit"),
                chains = 4, 
                iter = 5000,
                warmup = 500,
                control = list(max_treedepth = 20),
                cores = getOption("mc.cores", 4), 
                file = "output/fullmodtest")
```

```{r}
summary(fullmodtest)
```

```{r}
fullmodtest2 <- brm(formula = DEATH_EVENT ~ ., 
                data = heartt,
                family = bernoulli(link = "logit"),
                chains = 4, 
                iter = 2000,
                warmup = 500,
                control = list(max_treedepth = 15),
                cores = getOption("mc.cores", 4), 
                file = "output/fullmodtest2")
```

```{r}
summary(fullmodtest2)
```
```{r}
# Intercept-only model with test data
intmodtest <- brm(formula = DEATH_EVENT ~ 1, 
                data = heartt,
                family = bernoulli(link = "logit"),
                chains = 4, 
                iter = 2000,
                warmup = 500,
                control = list(max_treedepth = 15),
                cores = getOption("mc.cores", 4), 
                file = "output/intmodtest")
```

```{r}
summary(intmodtest)
```

```{r}
# Reduced model with test data
redmodtest <- brm(formula = DEATH_EVENT ~ ejection_fraction + serum_creatinine, 
                data = heartt,
                family = bernoulli(link = "logit"),
                chains = 4, 
                iter = 2000,
                warmup = 500,
                control = list(max_treedepth = 15),
                cores = getOption("mc.cores", 4), 
                file = "output/redmodtest")
```

```{r}
summary(redmodtest)
```

```{r}
# Full model test with 1,000 iterations
fullmodtest3 <- brm(formula = DEATH_EVENT ~ ., 
                data = heartt,
                family = bernoulli(link = "logit"),
                chains = 4, 
                iter = 1000,
                warmup = 500,
                control = list(max_treedepth = 15),
                cores = getOption("mc.cores", 4), 
                file = "output/fullmodtest3")
```

```{r}
summary(fullmodtest3)
```

```{r}
# Intercept-only model test with 1,000 iterations
intmodtest2 <- brm(formula = DEATH_EVENT ~ 1, 
                data = heartt,
                family = bernoulli(link = "logit"),
                chains = 4, 
                iter = 1000,
                warmup = 500,
                control = list(max_treedepth = 15),
                cores = getOption("mc.cores", 4), 
                file = "output/intmodtest2")
```

```{r}
summary(intmodtest2)
```

```{r}
# Reduced model test with 1,000 iterations
redmodtest2 <- brm(formula = DEATH_EVENT ~ ejection_fraction + serum_creatinine, 
                data = heartt,
                family = bernoulli(link = "logit"),
                chains = 4, 
                iter = 1000,
                warmup = 500,
                control = list(max_treedepth = 15),
                cores = getOption("mc.cores", 4), 
                file = "output/redmodtest2")
```

```{r}
summary(redmodtest2)
```

```{r}
# Full model
fullmod <- brm(formula = DEATH_EVENT ~ ., 
                data = heart,
                family = bernoulli(link = "logit"),
                chains = 4, 
                iter = 1000,
                warmup = 500,
               control = list(max_treedepth = 20),
                cores = getOption("mc.cores", 4), 
                file = "output/fullmod3")
```

```{r}
summary(fullmod)
```

```{r}
# Intercept-only model
intmod <- brm(formula = DEATH_EVENT ~ 1, 
                data = heart,
                family = bernoulli(link = "logit"),
                chains = 4, 
                iter = 1000,
                warmup = 500,
               control = list(max_treedepth = 20),
                cores = getOption("mc.cores", 4), 
                file = "output/intmod")
```

```{r}
summary(intmod)
```

```{r}
# Reduced model
redmod <- brm(formula = DEATH_EVENT ~ ejection_fraction + serum_creatinine, 
                data = heart,
                family = bernoulli(link = "logit"),
                chains = 4, 
                iter = 1000,
                warmup = 500,
               control = list(max_treedepth = 20),
                cores = getOption("mc.cores", 4), 
                file = "output/redmod")
```

```{r}
summary(redmod)
```

## In-Sample Model Checking

```{r}
# Posterior prediction for full model, intercept-only model, 
# and reduced model
ynewfull <- posterior_predict(fullmod)
ynewint <- posterior_predict(intmod)
ynewred <- posterior_predict(redmod)
```

```{r}
# Point estimates for all models
ytildefull <- apply(ynewfull, 2, mean)
ytildeint <- apply(ynewint, 2, mean)
ytildered <- apply(ynewred, 2, mean)
```

```{r}
# Residuals for all models
resfull <- heart$DEATH_EVENT - ytildefull
resint <- heart$DEATH_EVENT - ytildeint
resred <- heart$DEATH_EVENT - ytildered
```

```{r}
# Full model residuals plotted against age, binned
binnedplot(heart$age, resfull,
           xlab = "Age",
           main = "Full model residuals plotted against age")
```

```{r}
# Full model residuals plotted against ejection fraction, binned
binnedplot(heart$ejection_fraction, resfull,
           xlab = "Ejection fraction",
           main = "Full model residuals plotted against ejection fraction")
```

```{r}
# Full model residuals plotted against log serum creatinine, binned
binnedplot(log(heart$serum_creatinine), resfull,
           xlab = "Serum creatinine",
           main = "Full model residuals plotted against log serum creatinine")
```

```{r}
# Intercept-only model plotted against ejection fraction
binnedplot(heart$ejection_fraction, resint,
           xlab = "Ejection fraction",
           main = "Intercept-only model residuals plotted against ejection fraction")
```

```{r}
# Reduced model plotted against ejection fraction
binnedplot(heart$ejection_fraction, resred,
           xlab = "Ejection fraction",
           main = "Reduced model residuals plotted against ejection fraction")
```

```{r}
# Reduced model plotted against log serum creatinine
binnedplot(log(heart$serum_creatinine), resred,
           xlab = "Serum creatinine",
           main = "Reduced model residuals plotted against log serum creatinine")
```

All of the models underpredict death at low ejection fraction levels. So, a posterior predictive check should be done for ejection fraction at or below 25. 

```{r}
# Summary statistic function
summ_stat <- function(y, ejection_fraction)
  sum(y * (ejection_fraction <= quantile(ejection_fraction, 0.15))) / sum((ejection_fraction) <= quantile(ejection_fraction, 0.15))
```

```{r}
# Posterior predictive histogram for full model
tobs <- summ_stat(heart$DEATH_EVENT, heart$ejection_fraction)
tyrepfull <- apply(ynewfull, 1, summ_stat, heart$ejection_fraction)
hist(tyrepfull, main = "Posterior predictive histogram for full model")
abline(v = tobs)
```

```{r}
# Posterior predictive histogram for intercept-only model
tyrepint <- apply(ynewint, 1, summ_stat, heart$ejection_fraction)
hist(tyrepint, xlim = c(0.1, 0.7), main = "Posterior predictive histogram for intercept-only model")
abline(v = tobs)
```

```{r}
# Posterior predictive histogram for reduced model
tyrepred <- apply(ynewred, 1, summ_stat, heart$ejection_fraction)
hist(tyrepred, main = "Posterior predictive histogram for reduced model")
abline(v = tobs)
```

Based on these in-sample model checks, it seems that the full model actually performs a bit better than the reduced model, unlike the reference paper. This is a bit unsurprising, however, as the full model is more informed by the data than the reduced model. 

## Out-of-Sample Model Checking

## LOO Cross-Validation

```{r}
# Pareto k estimates, full model
loo_full <- loo(fullmod, save_psis = TRUE)
loo_full
```

```{r}
# Pareto k estimates, intercept-only model
loo_int <- loo(intmod, save_psis = TRUE)
loo_int
```

```{r}
# Pareto k estimates, reduced model
loo_red <- loo(redmod, save_psis = TRUE)
loo_red
```

```{r}
# PSIS plot, full model
plot(loo_full,
     diagnostic = c("k"),
     label_points = TRUE, 
     main = "PSIS diagnositc plot - full model")
```

```{r}
# PSIS plot, intercept-only model
plot(loo_int,
     diagnostic = c("k"),
     label_points = TRUE, 
     main = "PSIS diagnostic plot, intercept-only model")
```

```{r}
# PSIS plot, reduced model
plot(loo_red,
     diagnostic = c("k"),
     label_points = TRUE, 
     main = "PSIS diagnostic plot, reduced model")
```

## Test statistics

## T1: Proportion of Survival 

```{r}
# Full model
true_t1 <- mean(heart$DEATH_EVENT == 0)
hs_t1full <- sapply(1:nrow(ynewfull), function(x) mean(ynewfull[x, ] == 0))

t1_plotfull <- ggplot(data = as_tibble(hs_t1full), aes(value)) + 
  geom_histogram(aes(fill = "replicated")) + 
  geom_vline(aes(xintercept = true_t1, color = "observed"), lwd = 1.5) + 
  ggtitle("Survival proportion, full model") + 
  theme_bw(base_size = 16) + 
  scale_color_manual(name = "", values = c("observed" = "darkblue")) + 
  scale_fill_manual(name = "", values = c("replicated" = "lightblue"))
t1_plotfull
```

```{r}
mean(hs_t1full >= true_t1)
```

```{r}
# Intercept-only model
true_t1 <- mean(heart$DEATH_EVENT == 0)
hs_t1int <- sapply(1:nrow(ynewint), function(x) mean(ynewint[x, ] == 0))

t1_plotint <- ggplot(data = as_tibble(hs_t1int), aes(value)) + 
  geom_histogram(aes(fill = "replicated")) + 
  geom_vline(aes(xintercept = true_t1, color = "observed"), lwd = 1.5) + 
  ggtitle("Survival proportion, intercept-only model") + 
  theme_bw(base_size = 16) + 
  scale_color_manual(name = "", values = c("observed" = "darkblue")) + 
  scale_fill_manual(name = "", values = c("replicated" = "lightblue"))
t1_plotint
```

```{r}
mean(hs_t1int >= true_t1)
```

```{r}
# Reduced model
true_t1 <- mean(heart$DEATH_EVENT == 0)
hs_t1red <- sapply(1:nrow(ynewred), function(x) mean(ynewred[x, ] == 0))

t1_plotred <- ggplot(data = as_tibble(hs_t1red), aes(value)) + 
  geom_histogram(aes(fill = "replicated")) + 
  geom_vline(aes(xintercept = true_t1, color = "observed"), lwd = 1.5) + 
  ggtitle("Survival proportion, reduced model") + 
  theme_bw(base_size = 16) + 
  scale_color_manual(name = "", values = c("observed" = "darkblue")) + 
  scale_fill_manual(name = "", values = c("replicated" = "lightblue"))
t1_plotred
```

```{r}
mean(hs_t1red >= true_t1)
```

## T2: Proportion of Survival Among Patients with Ejection Fraction >40 Percent

```{r}
# Full model
ynew_ejecfull <- t(ynewfull) %>% 
  as_tibble() %>% 
  cbind(heart$ejection_fraction) %>% 
  filter(`heart$ejection_fraction` > 40)

true_t2 <- nrow(filter(heart, DEATH_EVENT == 0, ejection_fraction > 40)) / nrow(filter(heart, ejection_fraction > 40))
hs_t2full <- sapply(1:(ncol(ynew_ejecfull) - 1), FUN = function(x) mean(ynew_ejecfull[, x] == 0))

t2_plotfull <- ggplot(data = as_tibble(hs_t2full), aes(value)) + 
  geom_histogram(aes(fill = "replicated")) + 
  geom_vline(aes(xintercept = true_t2, color = "observed"), lwd = 1.5) + 
  ggtitle("Survival proportion among patients w/ ejection fraction >40%, full model") + 
  theme_bw(base_size = 12) + 
  scale_color_manual(name = "", values = c("observed" = "darkblue")) + 
  scale_fill_manual(name = "", values = c("replicated" = "lightblue")) 
t2_plotfull
```

```{r}
mean(hs_t2full >= true_t2)
```

```{r}
# Intercept-only model
ynew_ejecint <- t(ynewint) %>% 
  as_tibble() %>% 
  cbind(heart$ejection_fraction) %>% 
  filter(`heart$ejection_fraction` > 40)

true_t2 <- nrow(filter(heart, DEATH_EVENT == 0, ejection_fraction > 40)) / nrow(filter(heart, ejection_fraction > 40))
hs_t2int <- sapply(1:(ncol(ynew_ejecint) - 1), FUN = function(x) mean(ynew_ejecint[, x] == 0))

t2_plotint <- ggplot(data = as_tibble(hs_t2int), aes(value)) + 
  geom_histogram(aes(fill = "replicated")) + 
  geom_vline(aes(xintercept = true_t2, color = "observed"), lwd = 1.5) + 
  ggtitle("Survival proportion among patients w/ ejection fraction >40%, intercept-only model") + 
  theme_bw(base_size = 10) + 
  scale_color_manual(name = "", values = c("observed" = "darkblue")) + 
  scale_fill_manual(name = "", values = c("replicated" = "lightblue")) 
t2_plotint
```

```{r}
mean(hs_t2int >= true_t2)
```

```{r}
# Reduced model
ynew_ejecred <- t(ynewred) %>% 
  as_tibble() %>% 
  cbind(heart$ejection_fraction) %>% 
  filter(`heart$ejection_fraction` > 40)

true_t2 <- nrow(filter(heart, DEATH_EVENT == 0, ejection_fraction > 40)) / nrow(filter(heart, ejection_fraction > 40))
hs_t2red <- sapply(1:(ncol(ynew_ejecred) - 1), FUN = function(x) mean(ynew_ejecred[, x] == 0))

t2_plotred <- ggplot(data = as_tibble(hs_t2red), aes(value)) + 
  geom_histogram(aes(fill = "replicated")) + 
  geom_vline(aes(xintercept = true_t2, color = "observed"), lwd = 1.5) + 
  ggtitle("Survival proportion among patients w/ ejection fraction >40%, reduced model") + 
  theme_bw(base_size = 11) + 
  scale_color_manual(name = "", values = c("observed" = "darkblue")) + 
  scale_fill_manual(name = "", values = c("replicated" = "lightblue")) 
t2_plotred
```

```{r}
mean(hs_t2red >= true_t2)
```

## T3: Proportion of Survival Among Patients with Serum Creatinine <1.2

```{r}
# Full model
ynew_creafull <- t(ynewfull) %>% 
  as_tibble() %>% 
  cbind(heart$serum_creatinine) %>% 
  filter(`heart$serum_creatinine` < 1.2)

true_t3 <- nrow(filter(heart, DEATH_EVENT == 0, serum_creatinine < 1.2)) / nrow(filter(heart, serum_creatinine < 1.2))
hs_t3full <- sapply(1:(ncol(ynew_creafull) - 1), FUN = function(x) mean(ynew_creafull[, x] == 0))

t3_plotfull <- ggplot(data = as_tibble(hs_t3full), aes(value)) + 
  geom_histogram(aes(fill = "replicated")) + 
  geom_vline(aes(xintercept = true_t3, color = "observed"), lwd = 1.5) + 
  ggtitle("Survival proportion among patients with serum creatinine < 1.2, full model") + 
  theme_bw(base_size = 10) + 
  scale_color_manual(name = "", values = c("observed" = "darkblue")) + 
  scale_fill_manual(name = "", values = c("replicated" = "lightblue")) 
t3_plotfull
```

```{r}
mean(hs_t3full >= true_t3)
```

```{r}
# Intercept-only model
ynew_creaint <- t(ynewint) %>% 
  as_tibble() %>% 
  cbind(heart$serum_creatinine) %>% 
  filter(`heart$serum_creatinine` < 1.2)

true_t3 <- nrow(filter(heart, DEATH_EVENT == 0, serum_creatinine < 1.2)) / nrow(filter(heart, serum_creatinine < 1.2))
hs_t3int <- sapply(1:(ncol(ynew_creaint) - 1), FUN = function(x) mean(ynew_creaint[, x] == 0))

t3_plotint <- ggplot(data = as_tibble(hs_t3int), aes(value)) + 
  geom_histogram(aes(fill = "replicated")) + 
  geom_vline(aes(xintercept = true_t3, color = "observed"), lwd = 1.5) + 
  ggtitle("Survival proportion among patients with serum creatinine < 1.2, intercept-only model") + 
  theme_bw(base_size = 10) + 
  scale_color_manual(name = "", values = c("observed" = "darkblue")) + 
  scale_fill_manual(name = "", values = c("replicated" = "lightblue")) 
t3_plotint
```

```{r}
mean(hs_t3int >= true_t3)
```

```{r}
# Reduced model
ynew_creared <- t(ynewred) %>% 
  as_tibble() %>% 
  cbind(heart$serum_creatinine) %>% 
  filter(`heart$serum_creatinine` < 1.2)

true_t3 <- nrow(filter(heart, DEATH_EVENT == 0, serum_creatinine < 1.2)) / nrow(filter(heart, serum_creatinine < 1.2))
hs_t3red <- sapply(1:(ncol(ynew_creared) - 1), FUN = function(x) mean(ynew_creared[, x] == 0))

t3_plotred <- ggplot(data = as_tibble(hs_t3red), aes(value)) + 
  geom_histogram(aes(fill = "replicated")) + 
  geom_vline(aes(xintercept = true_t3, color = "observed"), lwd = 1.5) + 
  ggtitle("Survival proportion among patients with serum creatinine < 1.2, reduced model") + 
  theme_bw(base_size = 10) + 
  scale_color_manual(name = "", values = c("observed" = "darkblue")) + 
  scale_fill_manual(name = "", values = c("replicated" = "lightblue")) 
t3_plotred
```

```{r}
mean(hs_t3red >= true_t3)
```



